---
title: "Reproducible Research: Peer Assessment 1"
author: "Hannon Queiroz"
output: 
  html_document:
    keep_md: true
---
## Loading necessary libraries
```{r results='hide', message=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
```


## Loading and preprocessing the data
```{r}
activities <- fread('unzip -cq activity.zip')
activities$date <- as.Date(activities$date, "%Y-%m-%d")
print(head(activities))
```

## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day (ignoring NAs)
```{r}
steps_per_day <- activities %>%
    group_by(date) %>%
    summarise(
        total_steps = sum(steps, na.rm=TRUE)    
    )
print(head(steps_per_day))
```

2. Make a histogram of the total number of steps taken each day
```{r}
hist(steps_per_day$total_steps,
     main = "Histogram of total setps taken per day",
     xlab = "Total Steps",
     col = "blue")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
steps_mean <- mean(steps_per_day$total_steps, na.rm = TRUE)
steps_median <- median(steps_per_day$total_steps, na.rm = TRUE)
```
The mean of the total steps taken per day is `r steps_mean`.  
The mediean of the total steps taken per day is `r steps_median`.  


## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
steps_interval <- activities %>%
    group_by(interval) %>%
    summarise(
        number_steps = mean(steps, na.rm = TRUE)
    )
print(head(steps_interval))
ggplot(data = steps_interval, aes(x = interval, y = number_steps)) + 
    ggtitle("Average number of steps taken") +
    xlab("Interval") +
    ylab("Average number of steps") +
    geom_line()
```


2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_interval = arrange(steps_interval, desc(number_steps))[1,]
print(max_interval)
```

The 5-minute inteval that contains, on average, the maximum number of steps is inteval `r max_interval[1]`, with `r max_interval[2]` steps.


## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
number_na <- sum(is.na(activities$steps))
```
There are `r number_na` missing values in the dataset.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Let's fill in the missing values based on the mean for a given 5-minute interval:
```{r}
activities2 <- activities
activities2$steps <- apply(activities, 1, function(x){
    print(str(x))
    stop()
})
```





## Are there differences in activity patterns between weekdays and weekends?
